---
title: 'BIOST537: Problem Set 3'
date: "`r Sys.Date()`"
author: William Garcia
output: pdf_document
header-includes:
  # https://github.com/rstudio/rmarkdown/issues/646
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{xcolor}
  - \usepackage{booktabs}
  - \usepackage{caption} 
geometry: margin = 0.3 in
urlcolor: blue
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, message = FALSE, warning = FALSE,
                      fig.height = 4, fig.width = 6, fig.align = "center")
pacman::p_load(tidyverse) # additional packages can be included here
```

```{r prep, message=F, warning=F, echo=F}
# Load required libraries
library(readxl)
library(data.table)
library(knitr)
library(survival)
library(ggplot2)
library(survminer)
```

# Instructions
In all problems below, please do not simply provide a numerical answer: interpret the estimand(s) involved and summarize your findings in language appropriate for a general scientific audience.

# Problem 1

In the CCG803 study, 268 children in remission of acute lymphoblastic leukemia were recruited from a number of clinical institutions and randomized to one of two maintenance regimes. Patients in the control group were assigned to the standard of care --- chemotherapy with 6MP and methotrexate --- while patients in the treatment group were also given dactinomycin.

Dactinomycin administration (\texttt{rx}), age at baseline (\texttt{age}), white blood cell count (\texttt{wbc}), clinical institution (\texttt{institution}), observed follow-up time (\texttt{duration}) and relapse status at end of follow-up (\texttt{relapse}) were recorded for each patient. The dataset, called \texttt{ccg803.csv}, can be found on the Canvas website.


[(a)] Describe and fit a proportional hazards model to estimate the association between risk of relapse and treatment. Provide an estimate and 95\% confidence interval for the hazard ratio.
    
```{r 1a, message=F, warning=F, echo=F}
# Load the dataset
dt <- fread("ccg803.csv")

# Rx 1 = treated
dt[, rx := factor(rx, levels = c(0, 1), labels = c("control", "treated"))]

# (a) Unadjusted Cox model
cox_a <- coxph(Surv(duration, relapse) ~ rx, data = dt)
summary(cox_a)
```

\textcolor{violet}{I run a Cox proportional hazards model for the time until relapse as the outcome on treatment (rx). The estimand of interest is the hazard ratio (HR) comparing the instantaneous risk of relapse for patients receiving dactinomycin (treatment group) versus those receiving only the standard chemotherapy (control group). I estimate a HR of 0.74 with a 95\% confidence interval of (0.5497, 0.9917), which means that, on average, children in the treatment group experience about a 26\% lower hazard of relapse compared to controls.}

[(b)] Describe and fit a proportional hazards model to estimate the association between risk of relapse and treatment adjusting for white blood cell count and age. Provide an estimate and 95\% confidence interval for each of the resulting hazard ratios.

```{r 1b, message=F, warning=F, echo=F}
# (b) Cox model adjusted for age and white blood cell count (wbc)
cox_b <- coxph(Surv(duration, relapse) ~ rx + age + wbc, data = dt)
summary(cox_b)
```

\textcolor{violet}{I extend the model to adjust for age at baseline (age) and white blood cell count (wbc). The estimand of interest is the treatment effect (HR for rx) after accounting for age and wbc. A treatment HR of 0.74 (95\% CI: 0.55, 0.99), suggests that after adjusting for age and wbc, the treatment with dactinomycin might reduce the risk of relapse by about 26\%, though the upper limit of the confidence interval close to 1 and p-value of 0.0492 indicates that the estimate is just below a typical 5\% significance level.}

[(c)] Describe and fit a proportional hazards model to determine whether the association between risk of relapse and treatment adjusting for white blood cell count and age differs in subpopulations of patients defined by white blood cell count being either below 10,000 (\texttt{wbc} $<$ 100), or above or at 10,000 (\texttt{wbc} $\geq$ 100). Provide an estimate and 95\% confidence interval for the hazard ratio corresponding to each of these subpopulations. Perform a test of the hypothesis that this association does not differ in these subpopulations.

```{r 1c, message=F, warning=F, echo=F}
# (c) Testing effect modification by wbc

# Create a new variable for wbc group
dt[, wbc_group := ifelse(wbc < 100, "low", "high")]

# Fit a Cox model with an interaction rx * wbc_group
cox_c <- coxph(Surv(duration, relapse) ~ rx * wbc_group + age, data = dt)
summary(cox_c)

# Testing the interaction
cox_c_no_int <- coxph(Surv(duration, relapse) ~ rx + wbc_group + age, data = dt)
anova(cox_c_no_int, cox_c)
```

\textcolor{violet}{To explore whether the treatment effect differs by baseline white blood cell count, I dichotomize wbc into two groups (<100 or >=100) and include an interaction term between treatment and the wbc group in the Cox model with covariates. The likelihood ratio test with a non-significant 0.87 p-value (p > 0.05) indicate that there is no statistically significant evidence that the effect of dactinomycin differs between the low and high wbc groups.
}


[(d)] Describe and fit a proportional hazards model to estimate the association between risk of relapse and treatment adjusting for white blood cell count, age, and recruitment site. Provide an estimate and 95\% confidence interval for each of the resulting hazard ratios.

```{r 1d, message=F, warning=F, echo=F}
# (d) Cox model adjusted for age, wbc, and recruitment site (institution)

cox_d <- coxph(Surv(duration, relapse) ~ rx + age + wbc + factor(institution), data = dt)
summary(cox_d)
```

\textcolor{violet}{I further adjust the Cox model by including institution as a covariate (as a categorical variable). This model adjusts for treatment, age, wbc, and any differences between recruitment sites. The HR of 0.736 (95\% CI: 0.54, 1.00) after adding the recruitment site, suggest that the evidence for a protective effect of dactinomycin after accounting for potential site-specific differences is not significat at the 5\% significance level.}

[(e)] Based on the proportional hazards model you fitted in (b), display on a single graph estimates of the relapse-free survival curves for the subpopulation of:
    \begin{enumerate}
        \item[(i)] 5-year-old treated patients with \texttt{wbc} = 45;
        \item[(ii)] 5-year-old control patients with \texttt{wbc} = 45;
        \item[(iii)] 5-year-old treated patients with \texttt{wbc} = 210;
        \item[(iv)] 5-year-old control patients with \texttt{wbc} = 210.
    \end{enumerate}

```{r 1e, message=F, warning=F, echo=F}
# (e) Plotting estimated relapse-free survival curves

# Create a data.table with the profiles for prediction
newdata <- data.table(
  rx   = factor(c("treated", "control", "treated", "control"),
                levels = c("control", "treated")),
  age  = rep(5, 4),
  wbc  = c(45, 45, 210, 210)
)

# Obtain survival curves using model (b)
surv_fit <- survfit(cox_b, newdata = newdata)

# Create a categorical variable for legend labels
surv_dt <- data.table(
  time = rep(surv_fit$time, each = length(newdata$rx)),
  surv = as.vector(t(surv_fit$surv)),
  rx = rep(newdata$rx, times = length(surv_fit$time)),
  wbc = rep(newdata$wbc, times = length(surv_fit$time))
)


surv_dt[, group := paste(rx, ", WBC =", wbc)]

# Plot using ggplot2
ggplot(surv_dt, aes(x = time, y = surv, color = group, linetype = group)) +
  geom_step(size = 1) +
  labs(
    title = "Predicted survival curves",
    x = "Time",
    y = "Relapse-Free Survival Probability",
    color = "Patient Group",
    linetype = "Patient Group"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("red", "blue", "pink", "skyblue")) +
  scale_linetype_manual(values = c(1, 1, 2, 2)) +
  theme(legend.position = "bottom")
```

\textcolor{violet}{Using the Cox model which adjusted for age and wbc, I plot the relapse-free survival curves for patients with wbc = 45 and wbc = 210 by treatment arm. The survival curve for treated patients lie above that for control patients, indicating a higher probability of remaining relapse-free at any given time. Additionally, curves for different levels of wbc show that baseline white blood cell count might influence the risk of relapse.}

# Problem 2

In this problem, you will once more revisit the dataset on methadone maintenance programs for heroin addicts you studied in Problem Sets 1 and 2. Please refer to Problem Set 1 for a description of the relevant variables. The dataset, called \texttt{methadone.csv}, can be found on the Canvas website.

As indicated before, Caplehorn and Bell (1991) provided an analysis of these data, seeking to identify factors favoring the retention of subjects because maintenance is known to be effective only in patients who remain in treatment. They were particularly interested in determining whether daily dosage is related to the probability of retention, and whether other factors can be used to identify subjects at high risk for failing to be retained.

[(a)] Describe and fit a proportional hazards model to estimate the association between risk of exit from maintenance and methadone dosage adjusting for history of incarceration and clinic as regular predictors. Provide an estimate and 95\% confidence interval for each of the resulting hazard ratios.


```{r 2a, message=F, warning=F, echo=F}
# The dataset
dt <- fread("methadone.csv")

# (a) Cox Proportional Hazards Model
model_a <- coxph(Surv(time, event) ~ dose + prison + factor(clinic), data = dt)
summary(model_a)

# Extract hazard ratios and 95% confidence intervals
hr_a <- exp(coef(model_a))
ci_a <- exp(confint(model_a))
result_a <- data.table(Variable = names(hr_a), HR = hr_a,
                       Lower95CI = ci_a[,1], Upper95CI = ci_a[,2])
print("Results for Model 2a:")
print(result_a)
```

\textcolor{violet}{The Cox proportional hazards model estimates the association between the risk of exiting maintenance and methadone dosage, adjusting for clinic. For each additional mg/day of methadone, the hazard of exiting treatment is reduced by about 3.5\%. Individuals with a history of incarceration have about a 38\% higher hazard of exit. There is a statistically significant difference in the baseline risk of exit between the two clinics.}

[(b)] In the following questions, use a stratified proportional hazards model with clinic as a stratifying variable. Describe and fit a proportional hazards model to estimate the association between risk of exit from maintenance and methadone dosage adjusting for clinic via stratification as well as adjusting for history of previous incarceration. Provide an estimate and 95\% confidence interval for each of the resulting hazard ratios. Discuss the implication of using clinic as a stratifying variable. Has the interpretation of these hazard ratios changed relative to the model you fitted in (a)?

```{r 2b, message=F, warning=F, echo=F}
# (b) stratifying by Clinic
model_b <- coxph(Surv(time, event) ~ dose + prison + strata(clinic), data = dt)
summary(model_b)

# Extract hazard ratios (for dose and prison) and 95% confidence intervals
hr_b <- exp(coef(model_b))
ci_b <- exp(confint(model_b))
result_b <- data.table(Variable = names(hr_b), HR = hr_b,
                       Lower95CI = ci_b[,1], Upper95CI = ci_b[,2])
print("Results for Model 2b:")
print(result_b)
```

\textcolor{violet}{By stratifying on clinic, I allow the baseline hazard to differ between clinics so any potential confounding due to clinic-specific baseline risks is now removed.}

[(c)] Describe and fit a proportional hazards model to determine whether the association between methadone dosage and risk of exit from maintenance adjusting for history of incarceration and clinic is different in those with and without a history of incarceration. Provide an estimate and 95\% confidence interval for any summary that allows you to answer this question. Based on this model, provide an estimate and 95\% confidence interval for the hazard ratio comparing patients from a given clinic with a history of incarceration receiving 100 mg/day of methadone to patients from the same clinic without a history of incarceration receiving 40 mg/day of methadone.

```{r 2c, message=F, warning=F, echo=F}
# (c) Cox model Dose*Prison
model_c <- coxph(Surv(time, event) ~ dose * prison + factor(clinic), data = dt)
summary(model_c)

# Extract hazard ratios and 95% confidence intervals
hr_c <- exp(coef(model_c))
ci_c <- exp(confint(model_c))
result_c <- data.table(Variable = names(hr_c), HR = hr_c,
                       Lower95CI = ci_c[,1], Upper95CI = ci_c[,2])
print("Results for model 2c:")
print(result_c)

# Extract coefficients
coefs <- coef(model_c)
# Calculate the difference in linear predictor
diff_lp <- coefs["dose"] * (100 - 40) + coefs["prison"] + coefs["dose:prison"] * 100

# Hazard ratio for this comparison
HR_specific <- exp(diff_lp)

# calculate the standard error of the linear combination.
v <- c(100 - 40, 1, 100)
# Extract the relevant variance-covariance submatrix
vcov_sub <- vcov(model_c)[c("dose", "prison", "dose:prison"), c("dose", "prison", "dose:prison")]
var_lp <- as.numeric(t(v) %*% vcov_sub %*% v)
se_lp <- sqrt(var_lp)

# 95% CI on the log scale then exponentiate
lower_specific <- exp(diff_lp - 1.96 * se_lp)
upper_specific <- exp(diff_lp + 1.96 * se_lp)

result_specific <- data.table(specific_HR = HR_specific,
                              Lower95CI = lower_specific,
                              Upper95CI = upper_specific)
print("Results fo 100 mg/day, prison=1 vs. 40 mg/day, prison=0:")
print(result_specific)
```

\textcolor{violet}{The specific hazard ratio is 0.19. This result implies that, after accounting for the interaction, a patient with a history of incarceration on 100 mg/day has about a 81\% smaller hazard of exiting treatment compared to a non-incarcerated patient on 40 mg/day from the same clinic.}

# Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

